\chapter{Collaborating with other people}
\label{cha:collab}

As a completely decentralised tool, Mercurial doesn't impose any
policy on how people ought to work with each other.  However, if
you're new to distributed revision control, it helps to have some
tools and examples in mind when you're thinking about possible
workflow models.

\section{Collaboration models}

With a suitably flexible tool, making decisions about workflow is much
more of a social engineering challenge than a technical one.
Mercurial imposes few limitations on how you can structure the flow of
work in a project, so it's up to you and your group to set up and live
with a model that matches your own particular needs.

\subsection{Factors to keep in mind}

The most important aspect of any model that you must keep in mind is
how well it matches the needs and capabilities of the people who will
be using it.  This might seem self-evident; even so, you still can't
afford to forget it for a moment.

I once put together a workflow model that seemed to make perfect sense
to me, but that caused a considerable amount of consternation and
strife within my development team.  In spite of my attempts to explain
why we needed a complex set of branches, and how changes ought to flow
between them, a few team members revolted.  Even though they were
smart people, they didn't want to pay attention to the constraints we
were operating under, or face the consequences of those constraints in
the details of the model that I was advocating.

Don't sweep foreseeable social or technical problems under the rug.
Whatever scheme you put into effect, you should plan for mistakes and
problem scenarios.  Consider adding automated machinery to prevent, or
quickly recover from, trouble that you can anticipate.  As an example,
if you intend to have a branch with not-for-release changes in it,
you'd do well to think early about the possibility that someone might
accidentally merge those changes into a release branch.  You could
avoid this particular problem by writing a hook that prevents changes
from being merged from an inappropriate branch.

\subsection{Informal anarchy}

I wouldn't suggest an ``anything goes'' approach as something
sustainable, but it's a model that's easy to grasp, and it works
perfectly well in a few unusual situations.

As one example, many projects have a loose-knit group of collaborators
who rarely physically meet each other.  Some groups like to overcome
the isolation of working at a distance by organising occasional
``sprints''.  In a sprint, a number of people get together in a single
location (a company's conference room, a hotel meeting room, that kind
of place) and spend several days more or less locked in there, hacking
intensely on a handful of projects.

A sprint is the perfect place to use the \hgcmd{serve} command, since
\hgcmd{serve} does not requires any fancy server infrastructure.  You
can get started with \hgcmd{serve} in moments, by reading
section~\ref{sec:collab:serve} below.  Then simply tell the person
next to you that you're running a server, send the URL to them in an
instant message, and you immediately have a quick-turnaround way to
work together.  They can type your URL into their web browser and
quickly review your changes; or they can pull a bugfix from you and
verify it; or they can clone a branch containing a new feature and try
it out.

The charm, and the problem, with doing things in an ad hoc fashion
like this is that only people who know about your changes, and where
they are, can see them.  Such an informal approach simply doesn't
scale beyond a handful people, because each individual needs to know
about $n$ different repositories to pull from.

\subsection{A single central repository}

For smaller projects, migrating from a centralised revision control
tool, perhaps the easiest way to get started is to have changes flow
through a single shared central repository.  This is also the
most common ``building block'' for more ambitious workflow schemes.

Contributors start by cloning a copy of this repository.  They can
pull changes from it whenever they need to, and some (perhaps all)
developers have permission to push a change back when they're ready
for other people to see it.

Under this model, it can still sometimes make sense for people to pull
changes directly from each other, without going through the central
repository.  Consider a case in which I have a tentative bug fix, but
I am worried that if I were to publish it to the central repository,
it might subsequently break everyone else's trees as they pull it.  To
reduce the potential for damage, I can ask you to clone my repository
into a temporary repository of your own and test it.  This lets us put
off publishing the potentially unsafe change until it has had a little
testing.

In this kind of scenario, people usually use the \command{ssh}
protocol to securely push changes to the central repository, as
documented in section~\ref{sec:collab:ssh}.  It's also usual to
publish a read-only copy of the repository over HTTP using CGI, as in
section~\ref{sec:collab:cgi}.  Publishing over HTTP satisfies the
needs of people who don't have push access, and those who want to use
web browsers to browse the repository's history.

\subsection{The Linux kernel model}

The development of the Linux kernel has a shallow hierarchical
structure, surrounded by a cloud of apparent chaos.  Because most
Linux developers use \command{git}, a distributed revision control
tool with capabilities similar to Mercurial, it's useful to describe
the way work flows in that environment; if you like the ideas, the
approach translates well across tools.

At the center of the community sits Linus Torvalds, the creator of
Linux.  He publishes a single source repository that is considered the
``authoritative'' current tree by the entire developer community.
Anyone can clone Linus's tree, but he is very choosy about whose trees
he pulls from.

Linus has a number of ``trusted lieutenants''.  As a general rule, he
pulls whatever changes they publish, in most cases without even
reviewing those changes.  Some of those lieutenants are generally
agreed to be ``maintainers'', responsible for specific subsystems
within the kernel.  If a random kernel hacker wants to make a change
to a subsystem that they want to end up in Linus's tree, they must
find out who the subsystem's maintainer is, and ask that maintainer to
take their change.  If the maintainer reviews their changes and agrees
to take them, they'll pass them along to Linus in due course.

Individual lieutenants have their own approaches to reviewing,
accepting, and publishing changes; and for deciding when to feed them
to Linus.  In addition, there are several well known branches that
people use for different purposes.  For example, a few people maintain
``stable'' repositories of older versions of the kernel, to which they
apply critical fixes as needed.

This model has two notable features.  The first is that it's ``pull
only''.  You have to ask, convince, or beg another developer to take a
change from you, because there are no shared trees, and there's no way
to push changes into a tree that someone else controls.

The second is that it's based on reputation and acclaim.  If you're an
unknown, Linus will probably ignore changes from you without even
responding.  But a subsystem maintainer will probably review them, and
will likely take them if they pass their criteria for suitability.
The more ``good'' changes you contribute to a maintainer, the more
likely they are to trust your judgment and accept your changes.  If
you're well-known and maintain a long-lived branch for something Linus
hasn't yet accepted, people with similar interests may pull your
changes regularly to keep up with your work.

Reputation and acclaim don't necessarily cross subsystem or ``people''
boundaries.  If you're a respected but specialised storage hacker, and
you try to fix a networking bug, that change will receive a level of
scrutiny from a network maintainer comparable to a change from a
complete stranger.

To people who come from more orderly project backgrounds, the
comparatively chaotic Linux kernel development process often seems
completely insane.  It's subject to the whims of individuals; people
make sweeping changes whenever they deem it appropriate; and the pace
of development is astounding.  And yet Linux is a highly successful,
well-regarded piece of software.

\section{The technical side of sharing}

\subsection{Informal sharing with \hgcmd{serve}}
\label{sec:collab:serve}

Mercurial's \hgcmd{serve} command is wonderfully suited to small,
tight-knit, and fast-paced group environments.  It also provides a
great way to get a feel for using Mercurial commands over a network.

Run \hgcmd{serve} inside a repository, and in under a second it will
bring up a specialised HTTP server; this will accept connections from
any client, and serve up data for that repository until you terminate
it.  Anyone who knows the URL of the server you just started, and can
talk to your computer over the network, can then use a web browser or
Mercurial to read data from that repository.  A URL for a
\hgcmd{serve} instance running on a laptop is likely to look something
like \Verb|http://my-laptop.local:8000/|.

The \hgcmd{serve} command is \emph{not} a general-purpose web server.
It can do only two things:
\begin{itemize}
\item Allow people to browse the history of the repository it's
  serving, from their normal web browsers.
\item Speak Mercurial's wire protocol, so that people can
  \hgcmd{clone} or \hgcmd{pull} changes from that repository.
\end{itemize}
In particular, \hgcmd{serve} won't allow remote users to \emph{modify}
your repository.  It's intended for read-only use.

If you're getting started with Mercurial, there's nothing to prevent
you from using \hgcmd{serve} to serve up a repository on your own
computer, then use commands like \hgcmd{clone}, \hgcmd{incoming}, and
so on to talk to that server as if the repository was hosted remotely.
This can help you to quickly get acquainted with using commands on
network-hosted repositories.

\subsubsection{A few things to keep in mind}

Because it provides unauthenticated read access to all clients, you
should only use \hgcmd{serve} in an environment where you either don't
care, or have complete control over, who can access your network and
pull data from your repository.

The \hgcmd{serve} command knows nothing about any firewall software
you might have installed on your system or network.  It cannot detect
or control your firewall software.  If other people are unable to talk
to a running \hgcmd{serve} instance, the second thing you should do
(\emph{after} you make sure that they're using the correct URL) is
check your firewall configuration.

By default, \hgcmd{serve} listens for incoming connections on
port~8000.  If another process is already listening on the port you
want to use, you can specify a different port to listen on using the
\hgopt{serve}{-p} option.

Normally, when \hgcmd{serve} starts, it prints no output, which can be
a bit unnerving.  If you'd like to confirm that it is indeed running
correctly, and find out what URL you should send to your
collaborators, start it with the \hggopt{-v} option.

\subsection{Using \command{ssh} as a tunnel}
\label{sec:collab:ssh}

\subsection{Serving HTTP with a CGI script}
\label{sec:collab:cgi}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "00book"
%%% End: 
